---
title: "Acquiring, preparing, and visualizing data"
author: "J.M.T. Roos"
date: "Last updated 7 May 2016"
autosize: false
css: "../css/slides.css"
output: 
  revealjs::revealjs_presentation:
    self_contained: true
    theme: simple
    highlight: NULL
    transition: slide
    slide_level: 1
    reveal_options:
      controls: false
      margin: .05
    pandoc_args:
      [
      "--variable", "slideNumber='c/t'",
      "--variable", "transitionSpeed=default",
      "--variable", "viewDistance=5"
      ]
---



```{r echo=FALSE, include=FALSE}
knitr::opts_chunk$set(tidy = TRUE, tidy.opts=list(blank=FALSE, width.cutoff=50), cache=TRUE)
knitr::knit_hooks$set(
source = function(x,options){
  if( options$engine == 'R' ) {
    # format R code
    x = highr::hilight( x, format = 'html' )
  } else if( options$engine == 'bash' ) {
    # format bash code
    x = paste0( '<span class="hl std">$</span> ', unlist( stringr::str_split( x, '\\n' ) ), '\n', collapse = '' )
  }
  # for (i in names(options)) {
  #   message( paste( i, options[[i]] ) )
  # }
  x = paste(x, collapse = "\n")
  sprintf("<div class=\"%s\"><pre class=\"%s %s\"><code class=\"%s %s\">%s</code></pre></div>\n", 'sourceCode', 'sourceCode', tolower(options$engine), 'sourceCode', tolower(options$engine), x)
}
)
```


Review
==================================
* In class
    * git --- cloning, committing, pushing
    * R Markdown --- mixing text with R code using `` ```{r} `` blocks
* During the week
    * Reviewing/learning the R programming language
    * Installing packages

```{r echo=FALSE, message=FALSE}
library(tidyr)
library(dplyr)
library(reshape2)
library(ggplot2)
```

```{r echo=FALSE}
test_data <- NULL
local( {
  relabel_factors <- function( z ) { eval(parse(text=paste0('c(',paste0(z,1:4,'=',1:4,collapse = ','),')') )) }

  test_data <<- datasets::anscombe %>% melt(id.vars=paste0('y',1:4), value.name = 'num.correct', variable.name = 'x') %>% tbl_df %>% melt(id.vars = c('x','num.correct'), value.name = 'duration', variable.name = 'y') %>% tbl_df %>% mutate( x = plyr::revalue(x,relabel_factors('x')),  y = plyr::revalue(y,relabel_factors('y'))) %>% filter( x == y ) %>% group_by( x ) %>% mutate( respondent = factor(1:n()) ) %>% select( round = x, respondent, num.correct, duration ) %>% ungroup
} )
```


Applied problem: Merging samples
=================================
Repeated measures for 11 individuals, mean (sd)

|Round |Duration |Number Correct |
|:----:|:-------:|:----------:|
|1     |7.5      |9.0         |
|      |(2.0)    |(3.3)       |
|2     |7.5      |9.0         |
|      |(2.0)    |(3.3)       |
|3     |7.5      |9.0         |
|      |(2.0)    |(3.3)       |
|4     |7.5      |9.0         |
|      |(2.0)    |(3.3)       |


Regression of `duration` on `num.correct` ($\times$ 4)

|Variable  |Estimate|SE|
|:---------|-------:|-:|            
|Intercept  | 3.00   |  1.12   |
|num.correct| 0.50|   0.12   |

***

<div class='center'>
```{r echo=FALSE}
test_data %>% ggplot( aes( x = duration, y = num.correct ) ) + geom_point() + facet_wrap( ~round, labeller = label_both )
```
</div>

***

<div class='center'>
```{r echo=FALSE}
test_data %>% ggplot( aes( x = duration, y = num.correct ) ) + geom_point() + facet_wrap( ~round, labeller = label_both ) + geom_smooth(method="lm",fullrange=TRUE)
```
</div>

Remember...
========================
<div class="center">
## Always look at the data first
</div>


Today
==================================
* Data visualization
* Acquiring (tabular) data
* Tidying and summarizing data


ggplot2
===================================
* Plotting package in R intended to replace the core plotting routines
* Based on the concept of a **grammar of graphics**
    * Plots are constructed from parts much as sentences are constructed from nouns, verbs, etc. 
    * Not all arrangements of words lead to comprehensible sentences -- the same is true for plots, and the grammar helps you avoid (visual) nonsense
    * Leads to a modularity of design, making it easy for programmers to extend
* Sensible and aesthetically pleasing default settings
    * Informed by what we know about visual perception and cognition

What is a graph?
===============
> A visual display that illustrates one or more relationships among numbers...a shorthand means of presenting information that would take many more words and numbers to describe.
> 
> ---Stephen M. Kosslyn. Graph Design for the Eye and Mind. Oxford University Press, 2006

It depends on the goal:

* A tool for discovery --- gain an overview of, convey the scale and complexity of, or facilitate an exploration of **data** (*dataviz*)
* A tool for communication --- help you to help others understand, tell a story about, or stimulate interest in a **problem or solution** (*infographics*)

At a minimum...
================
* Graphs are for comparing quantities
    * Always ask yourself: "Comparing what?"
    * Insist on this comparison being obvious to the viewer (and yourself)
* Graphs should answer a central question
    * Both question and answer must be clear
    * Use the title, caption, and other labels to highlight both

Psychological principles (Kosslyn, 2006)
========================================
* Connecting with the audience 
* Directing and holding attention 
* Promoting understanding and memory


Connecting with the audience
==========================
1. **Relevance**
    * Not too much or too little information
    * Present information that reflects the message you want to convey
    * Don’t present extraneous information 
2. **Appropriate knowledge**
    * Prior knowledge must be sufficient to understand the graph
    * If you assume too much prior knowledge, viewers will be confused
    * If you violate norms, viewers will be confused

Directing and holding attention 
================================
3. **Salience**
    * Attention is drawn to large perceptible differences
    * The most visually striking aspect receives the most attention
    * Annotations help direct viewers' attention
4. **Discriminability**
    * Properties must differ enough to be noticed
    * Defaults in `ggplot2` do much of this work for you
5. **Organization**
    * Groups of elements are seen and remembered as a whole

Understanding and memory 
=========================
6. **Compatibility**
    * Form should be aligned with meaning
    * Lines express continuous change, bars discrete quantities
    * More = more (higher, better, bigger, etc.)
7. **Informative changes**
    * Changes in properties should carry information
    * ...and vice versa
8. **Capacity limitations**
    * If too much information is presented, none is remembered
    * Four chunks in working memory
    * Graph designers err on the side of presenting too much, graph readers err on the side of paying too little attention



ggplot2's grammar
===================
* Decomposes graphs into basic parts
* Sets rules for interactions among those parts 
* Helps us stay out of trouble


ggplot2's grammar
===================
* **Default** values for **Data** and **Mapping** available to all layers
* **Layers** --- one or more, each with the following:
    * **Data** (overriding the default) --- a `data.frame`
    * **Mapping** (overriding the default) of columns to **Aesthetics**
    * **Geometry** specifying what to draw
    * **Statistic** specifying how to transform the data before drawing
    * **Position** specifying how to arrange items
* **Facet** specification for generating subplots
* **Scales** specifying how to translate the data to lengths, colors, sizes, etc. in the graph
* **Coordinates** which is the default (Cartesian) 99% of the time, so ignore for now

Layers
======
* Layers contain everything we see, often showing different views of the same data

***

<div class="center">
```{r echo=FALSE}
ggplot( data=anscombe, aes( x=x1, y = y1 ) ) + geom_point() + labs(title = 'Layer 1: data as points', x = 'x', y = 'y' )
```
</div>

***

<div class="center">
```{r echo=FALSE}
ggplot( data=anscombe, aes( x=x1, y = y1 ) ) + geom_smooth(method="lm") + labs(title = 'Layer 2: regression line', x = 'x', y = 'y' )
```
</div>

***

<div class="center">
```{r echo=FALSE}
ggplot( data=anscombe, aes( x=x1, y = y1 ) ) + geom_point() + geom_smooth(method="lm") + labs(title = 'Layer 1 and 2', x = 'x', y = 'y' )
```
</div>

Test data
=========
```{r}
test_data
```

Defaults
========
* Specify the defaults first
* Most graphs use a single set of **data** (`data.frame`) for every layer
* Most graphs use a single set of **mapping**s between columns and aesthetics

```{r}
my_plot <- ggplot( data = test_data, mapping = aes( x = duration, y = num.correct ) )
```

* `aes()` is used to create a list of `aes`thetic mappings
    * `x` refers to the graph's x-axis, `y` to the y-axis
    * `duration` $\rightarrow$ x-axis
    * `num.correct` $\rightarrow$ y-axis
* `my_plot` now represents a `ggplot` object set to our defaults
* You don't need to name the arguments; `data` comes first, `mapping` comes second

```{r}
my_plot <- ggplot( test_data, aes( x = duration, y = num.correct ) )
```

An empty plot
=============
* Defaults by themselves do nothing


```{r fig.height=3}
print( my_plot )
```

* By default, we get an "empty" plot
* To see something, we need to specify a layer

Adding a layer
===============
* Use the `+` operator to combine ggplot elements

```{r fig.height=3}
my_plot + geom_point()
```

* Usually you do not need the `print()` call, so the following two lines are equivalent:
```{r eval=FALSE}
my_plot + geom_point()
print( my_plot + geom_point() )
```

Each layer has a geometry
=========================
```{r fig.height=2.5}
my_plot + geom_point()
my_plot + geom_line()
```

***

<div class='center'>
```{r}
my_plot + geom_point() + geom_line()
```
</div>

Each layer has a statistic
==========================
* Usually the statistic is the `identity` function, $$f(x)=x$$ That is, the data are left unchanged
* The default statistic for `geom_point` and `geom_line` is `identity` so these plots show the data as is
* The default statistic for `geom_histogram` is a binning function (called `stat_bin`)

```{r fig.height=2}
ggplot( test_data, aes( x = duration ) ) + geom_histogram( binwidth = 2 )
```

***

<div style="font-size: 80%;">
```{r echo=FALSE, fig.show='hide'}
local({
  g <- ggplot( test_data, aes( x = duration ) ) + geom_histogram( binwidth = 2 )
  p <- print(g)
  print(test_data)
  print( p$data[[1]] %>% select(x,y) %>% tbl_df )
})
```
</div>

Geoms and statistics
================================================
* Each geom/statistic has a default statistic/geom


|Item    |   Default stat/geom|
|:-------|:-------------------|
|`geom_point`|`stat_identity` ($f(x)=x$)|
|`geom_line`|`stat_identity` ($f(x)=x$)|
|`geom_histogram`|`stat_bin` (binning) |
|`geom_smooth`|`stat_smooth` (regression) |
|`stat_smooth`|`geom_smooth` (line + ribbon) |
|`stat_bin`|`geom_bar` (vertical bars)|
|`stat_identity`|`geom_point` (dots)|

* Hence, these produce the same output:
```{r eval=FALSE}
ggplot( test_data, aes(x=duration) ) + stat_bin(binwidth=1)
ggplot( test_data, aes(x=duration) ) + geom_histogram(binwidth=1)
```

Data versus statistics
=====================
* Be sure you understand: "Does this layer contain data or statistics?"
* When in doubt, prefer data to statistics:
* Example: A scatter plot of observations conveys more information than a box plot showing quantiles

<div class="left">
```{r fig.height=3.5, fig.width=4.5, tidy=FALSE}
ggplot(test_data, aes(x = round, 
  y = duration)) + geom_point()
```
</div>
<div class="right">
```{r fig.height=3.5, fig.width=4.5, tidy=FALSE}
ggplot(test_data, aes(x = round, 
  y = duration)) + geom_boxplot()
```
</div>

Aesthetics
==========
* Each geometry interacts with one or more aesthetics

|Item    |   Required |  Optional |
|:-------|:-------------------|:----------------------|
|`geom_point`|`x`,&nbsp;`y`|`alpha`,&nbsp;`colour`,&nbsp;`fill`,&nbsp;`shape`,&nbsp;`size`,&nbsp;`stroke`|
|`geom_line`|`x`,&nbsp;`y`|`alpha`,&nbsp;`colour`,&nbsp;`linetype`,&nbsp;`size`|
|`geom_pointrange`|`x`,&nbsp;`ymax`,&nbsp;`ymin`|`alpha`,&nbsp;`colour`,&nbsp;`linetype`,&nbsp;`size`|

* You can either map data to an aesthetic, or set it explicitly

<div class='left'>
```{r fig.height=3, fig.width=4.5, tidy=FALSE}
my_plot + geom_point(
  mapping = aes(colour = round))
```
</div>
<div class='right'>
```{r fig.height=3, fig.width=4.5, tidy=FALSE}
my_plot + geom_point(
  colour="red")
```
</div>


Position
==========
* Each layer also has a position specification
* The default is again `identity` meaning don't do anything special
* Examples: bars can be positioned with `stack` or `dodge`

```{r}
g <- ggplot(test_data, aes(x = num.correct, fill = round ))
```
<div class="left">
```{r fig.height=3.5, fig.width=4.5, tidy=FALSE}
g + stat_bin(binwidth=4,
             position='stack')
```
</div>
<div class="right">
```{r fig.height=3.5, fig.width=4.5, tidy=FALSE}
g + stat_bin(binwidth=4,
             position='dodge')
```
</div>


Practice with layers
=====================
* Work with a neighbor
* First discuss the task, then one of you does the typing
* Discuss what you are doing as you write code
* Write your code in an empty **File > New File... > R Script** and execute each line using `Cmd-Enter`
* Use the data set called `mpg` which is included in the `ggplot2` package


Data
=======
```{r tidy=FALSE, eval=FALSE}
library(ggplot2)
?mpg
```

<div style="font-size: 60%;">
**Usage**
```
mpg
```

<div class="left">
**Format**

A data frame with 234 rows and 11 variables

* manufacturer.
* model.
* displ. engine displacement, in litres
* year.
* cyl. number of cylinders
* trans. type of transmission
* drv. f = front-wheel drive, r = rear wheel drive, 4 = 4wd
* cty. city miles per gallon
* hwy. highway miles per gallon
* fl.
* class.
</div>

<div class="right">
**Description**

This dataset contains a subset of the fuel economy data that the EPA makes
available on <a href = 'http://fueleconomy.gov'>http://fueleconomy.gov</a>. It contains only models which
had a new release every year between 1999 and 2008 - this was used as a
proxy for the popularity of the car.
</div>
</div>

***

```{r}
mpg
```


Task 0 (Example)
================
* Create a plot with 1 layer:
    * `x` mapped to `cty`
    * `y` mapped to `hwy`
    * `point` geometry
    * `identity` stat
    * `identity` position

Task 1
=================
* Create a plot with 1 layer:
    * `x` mapped to `displ`
    * `y` mapped to `hwy`
    * `colour` mapped to `trans`
    * geometry `point`
        * default stat and position for `point`


***

```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour = class)) + geom_point()
```


Task 2
======
* Defaults for all layers:
    * `x` $\rightarrow$ `displ`
    * `y` $\rightarrow$ `hwy`
* Layer 1: 
    * geom `point`
    * default stat and position
    * point colour is set to `'red'`
* Layer 2:
    * geom `smooth`
    * default stat and position

***

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point(colour='red') + geom_smooth()
```

Task 3
======
* Defaults for all layers:
    * `x` $\rightarrow$ `displ`
    * `y` $\rightarrow$ `hwy`
    * `colour` $\rightarrow$ `drv`
* Layer 1: 
    * geom `point`
    * default stat and position
* Layer 2:
    * geom `smooth`
    * default stat and position

***

```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour=drv)) + geom_point() + geom_smooth()
```

Task 4
======
* Defaults for all layers:
    * `x` $\rightarrow$ `displ`
    * `y` $\rightarrow$ `hwy`
    * `colour` $\rightarrow$ `drv`
* Layer 1: 
    * geom `point`
    * default stat and position
* Layer 2:
    * geom `smooth`
    * default stat and position
    * pass in `method='lm'` to get linear (rather than LOESS) regression
    * pass in `se=FALSE` to suppress confidence bands

***

```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour=drv)) + geom_point() + geom_smooth(method="lm", se=FALSE)
```

Task 5
======
* Defaults for all layers:
    * `x` $\rightarrow$ `displ`
    * `y` $\rightarrow$ `hwy`
    * `colour` $\rightarrow$ `cyl` (as discrete values)
* Layer 1: 
    * geom `point`
    * default stat and position
* Layer 2:
    * geom `smooth`
    * default stat and position

***

**Incorrect**
```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour=cyl)) + geom_point() + geom_smooth()
```

***

**Correct**
```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour=factor(cyl))) + geom_point() + geom_smooth()
```

Facets
=======

Something here...

Scales
======

Something here...


More reading
============
* [39 studies about human perception in 30 minutes](https://medium.com/@kennelliott/39-studies-about-human-perception-in-30-minutes-4728f9e31a73) is a nice review of what we know about how perception of data visualizations
* See the [ggplot2 documentation](http://docs.ggplot2.org/current/) for a visual summary of the available geometries, list of stats, and more; as well as detailed documentation
* [Chapter 3 of R for Data Science](http://r4ds.had.co.nz/data-visualisation.html) has a very nice introduction to ggplot2 that follows a similar flow to what we covered today



dplyr
=====




tidyr
=====
```{r eval=FALSE}
devtools::install_github('garrettgman/DSR')
```

